---
title: "Dashboard User Guide"
author: "Your Name"
date: "3 November 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dashboard User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

The `healthburdenr` interactive dashboard provides a comprehensive interface for exploring healthcare-associated infection (HAI) burden data. This guide walks you through each feature of the dashboard and explains how to interpret the visualizations.

## Launching the Dashboard

Load the package and launch the dashboard:
```{r launch, eval=FALSE}
library(healthburdenr)

# Launch with default settings
run_app()

# Launch on specific port (if default is already in use)
run_app(port = 3838)

# Launch without auto-opening browser
run_app(launch.browser = FALSE)
```

The dashboard will open in your default web browser. If it doesn't open automatically, look for a message in the R console showing the local URL (typically `http://127.0.0.1:####`).

## Dashboard Overview

The dashboard consists of three main tabs accessible from the left sidebar:

1. **Overview** - Survey sample distribution and HAI prevalence
2. **Population Estimates** - Annual burden estimates with visualizations
3. **Stratified Analysis** - Demographic breakdowns by age and sex

Each tab is self-contained with its own controls, visualizations, and interpretation guides.

---

## Tab 1: Overview

### Purpose

The Overview tab presents survey sample distribution data from Point Prevalence Surveys (PPS), showing:

- How many patients were surveyed in each region
- How many patients had healthcare-associated infections
- The distribution of specific infection types

### Key Features

#### Value Boxes

Three summary statistics provide a quick overview:

- **Total Survey Patients**: Combined total across Germany and EU
- **Total HAI Patients**: Total patients with at least one HAI
- **Infection Types Tracked**: Number of infection categories (5: HAP, SSI, BSI, UTI, CDI)

These boxes give you an at-a-glance understanding of the survey scale.

#### Information Boxes

Two collapsible boxes provide essential background:

- **About BHAI & Healthcare-Associated Infections**: Explains methodology and burden metrics
- **Understanding the Data Fields**: Definitions of key terms and infection types

Expand these if you're new to HAI surveillance or need clarification on terminology.

#### Interactive Treemap Visualization

The main feature of this tab is a hierarchical treemap showing:

**Structure:**

- **Top level**: Countries (Germany, European Union)
- **Second level**: HAI status (patients with HAI vs. unaffected)
- **Bottom level**: Specific infection types (HAP, SSI, BSI, UTI, CDI)

**How to interact:**

- **Hover**: Move your mouse over any rectangle to see exact numbers and percentages
- **Click**: Click on rectangles to zoom in and explore that segment
- **Breadcrumb**: Use the breadcrumb trail at the top to navigate back

**Color coding:**

- Blue shades: Unaffected patients
- Red/warm tones: Patients with HAI
- Distinct colors: Each infection type

**What to look for:**

- **Relative sizes**: Larger boxes = more patients
- **Prevalence**: Compare "HAI" vs. "Unaffected" sections to see overall infection rates
- **Common infections**: Identify which infections affect the most patients
- **Country differences**: Compare patterns between Germany and EU

### Interpretation Tips

1. **Prevalence patterns**: If the "HAI" section is much smaller than "Unaffected," prevalence is low (which is good)

2. **Infection distribution**: Within the HAI section, the largest rectangles show which infections are most commonly found

3. **Geographic variation**: Differences between Germany and EU may reflect:

   - Different healthcare systems
   - Different patient populations
   - Survey methodology variations

4. **Clinical implications**: Common infections (large boxes) may need widespread prevention efforts, while less common infections might require targeted approaches

---

## Tab 2: Population Estimates

### Purpose

This tab presents annual burden estimates derived from BHAI modeling, showing:

- Total cases, deaths, and DALYs by infection type
- Confidence intervals reflecting statistical uncertainty
- Interactive visualizations for exploring relationships

### Key Features

#### Summary Value Boxes

Four boxes display aggregate burden from the "ALL" row (sum across all infections):

- **Total HAI Cases**: Annual infection episodes with 95% CI
- **Total HAI Deaths**: Annual mortality with 95% CI
- **Overall Case-Fatality Rate**: Percentage of cases resulting in death
- **Total DALYs**: Combined burden of mortality and morbidity with 95% CI

These provide a high-level view of total HAI impact in the selected country.

#### Control Panel

**Country/Region Filter:**

- Select "Germany" or "European Union"
- All outputs update immediately when changed

**Data Notes:**

- Quick reference for understanding value displays
- Reminds you about confidence intervals and the "ALL" row

#### Tabbed Outputs

The main panel has two tabs:

**Tab 2a: Data Table**

A comprehensive table showing all burden metrics:

**Features:**

- Sortable columns (click headers)
- Searchable (use search box)
- Download buttons (Copy or CSV)
- Color-coded CFR (Case-Fatality Rate):

  - Green (<3%): Lower mortality
  - Yellow (3-10%): Moderate mortality
  - Red (>10%): Higher mortality

- Gray highlighting for "ALL" row (aggregate)

**Columns:**

- HAI: Infection type
- Cases: Point estimate with 95% CI
- Deaths: Point estimate with 95% CI
- CFR (%): Case-fatality rate
- DALYs, YLL, YLD: Population health loss metrics with CIs

**How to use:**

1. Sort by deaths to identify deadliest infections
2. Sort by cases to identify most common infections
3. Compare CIs to understand uncertainty
4. Download data for external analysis or reporting

**Tab 2b: Bubble Chart**

An interactive scatter plot showing relationships between burden metrics:

**Axes:**

- **X-axis**: Number of cases (annual incidence)
- **Y-axis**: Number of deaths (annual mortality)
- **Bubble size**: DALY burden (larger = more population health loss)
- **Color**: Infection type (consistent with Overview tab)

**How to interact:**

- **Hover**: See exact values, including confidence intervals and CFR
- **Zoom**: Use plotly toolbar to zoom and pan
- **Download**: Save as PNG using camera icon in toolbar

**What to look for:**

1. **High burden infections**: Large bubbles in the top-right

   - High incidence AND high mortality
   - Priority for prevention efforts

2. **High mortality rate**: Infections high on Y-axis relative to X-axis

   - Fewer cases but more lethal
   - May need specialized treatment protocols

3. **High incidence**: Infections far to the right

   - Many cases but lower mortality
   - May still have high DALY burden due to volume

4. **Overall burden**: Bubble size shows combined impact

   - Large bubbles are high priority regardless of position
   - Consider both frequency and severity

**Example interpretation:**

If BSI (Bloodstream Infection) appears as a large bubble in the top-right:

- High number of cases 
- High number of deaths
- Large DALY burden
- **Conclusion**: BSI is a high-priority infection requiring aggressive prevention and treatment strategies

#### Interpretation Box

A gray box provides tips for understanding the visualization:

- Position interpretation
- Bubble size meaning
- Prevention priorities
- How to use hover information

### Best Practices for This Tab

1. **Start with the table**: Get familiar with the numbers and ranges

2. **Use the bubble chart for patterns**: Visual relationships are easier to see in the chart

3. **Consider both metrics**: High incidence and high mortality both matter, but differently

4. **Pay attention to CIs**: Wide confidence intervals mean more uncertainty

5. **Compare countries**: Switch between Germany and EU to see differences

6. **Export for reports**: Download the table for inclusion in presentations or papers

---

## Tab 3: Stratified Analysis

### Purpose

This tab breaks down burden by demographic characteristics:

- Age groups
- Biological sex (Male, Female)
- Infection types

This allows identification of vulnerable populations and targeting of prevention efforts.

### Layout

**Left Column:**

- Control panel (filters)
- Interpretation guide (collapsible)
- Glossary (collapsible)

**Right Column:**

- Diverging bar chart visualization

### Control Panel

**Country/Region:**

- Select Germany or European Union
- Updates all outputs

**Metric to Display:**

- **Cases**: Number of infection episodes
- **Deaths**: Number of deaths
- **DALYs**: Disability-Adjusted Life Years

**Select Infections:**

- Check/uncheck infections to display
- **Tip**: Select 1-3 infections for best readability
- More than 3 can make the chart crowded

**Age Group Filter:**

- Select specific age groups or "All"
- Use this to focus on particular age ranges
- Multiple selection allowed

### Diverging Bar Chart

**Structure:**

- **Left panel (F)**: Female burden, bars extend right
- **Right panel (M)**: Male burden, bars extend right
- **Y-axis**: Age groups (youngest at bottom, oldest at top)
- **X-axis**: Burden metric (absolute scale shown)
- **Colors**: Different color for each infection type
- **Error bars**: Black horizontal lines showing 95% confidence intervals

**How to read:**

1. **Bar length**: Longer bars = higher burden in that age-sex group

2. **Side-by-side comparison**: Direct visual comparison of male vs. female

3. **Grouped bars**: When multiple infections selected, they appear side-by-side within each age group

4. **Error bars**: Wider error bars = more uncertainty

   - Common in very young (0-1) and very old (85+) groups due to smaller samples

**What to look for:**

1. **Age patterns**:

   - Which age groups have the longest bars?
   - Is burden concentrated in elderly (typical)?
   - Are there unexpected patterns in younger groups?

2. **Sex differences**:

   - Are bars consistently longer on one side?
   - Do patterns differ by infection type?
   - Are sex differences larger in certain age groups?

3. **Infection-specific patterns**:

   - Do different infections show different age distributions?
   - Which infections primarily affect elderly?
   - Which infections are more evenly distributed?

4. **Uncertainty**:

   - Which groups have wide confidence intervals?
   - Where is more data/surveillance needed?

**Example interpretation:**

If you see:

- Very long bars for ages 65-84 and 85+
- Slightly longer bars on the male side
- Wide error bars for age 0-1

**Conclusion**:

- Elderly populations (65+) carry the highest burden (typical finding)
- Males may have slightly higher burden than females (could be due to healthcare utilization or biological factors)
- Infant data has high uncertainty (may need more surveillance in this group)

### Interpretation Guide

Located below the control panel, this box provides:

- Key insights to look for
- Clinical implications
- Statistical notes
- Tips for interpretation

**Expand this** if you're unsure how to interpret patterns you see in the chart.

### Glossary (Collapsible)

Provides quick definitions of technical terms:

- Stratification
- Age groups
- Stratum
- Point estimates
- Confidence intervals
- Error bars
- Faceting

---

## General Dashboard Features

### Navigation

- **Sidebar menu**: Click tab names to switch between Overview, Population Estimates, and Stratified Analysis
- **Active tab**: Highlighted in blue
- **Collapsible boxes**: Click title bars to expand/collapse information boxes
- **Breadcrumbs**: In treemap, use breadcrumb trail to navigate hierarchy

### Interactivity

All visualizations are interactive:

- **Hover**: See detailed information
- **Click**: Interact with elements (zoom in treemap, etc.)
- **Zoom/Pan**: Use plotly toolbar on charts
- **Download**: Export visualizations and data

### Performance Tips

For the best experience:

1. **Use modern browsers**: Chrome, Firefox, Safari, or Edge
2. **Close unused tabs**: Reduces memory usage
3. **Select fewer infections**: In stratified analysis, limit to 2-3 infections
4. **Wait for rendering**: Large visualizations may take a moment to render

---

## Common Questions

### Why do some estimates have wide confidence intervals?

Wide confidence intervals occur when:

- Sample sizes are small in specific strata (e.g., rare infections, young age groups)
- Variability in the underlying data is high
- The estimation model has inherent uncertainty

This is normal and indicates areas where more surveillance data could improve precision.

### What does the "`ALL`" row mean?

The "`ALL`" row in the Population Estimates tab represents the sum of the five infection types (HAP, SSI, BSI, UTI, CDI). It's not a separate category, but rather the total HAI burden across all tracked infections.

*Important:* The "`ALL`" row does NOT appear in stratified data, since stratification is done for individual infections.

### Can I export the data?

Yes! Multiple export options:

Data tables: Use "`Copy`" or "`CSV`" buttons

Visualizations: Use the camera icon in the plotly toolbar to save as PNG

## Best Practices for Dashboard Use

### 1. Start Broad, Then Narrow

- Begin with Overview for context
- Move to Population Estimates for overall burden
- Use Stratified Analysis to identify specific high-risk groups

### 2. Always Consider Uncertainty

- Look at confidence intervals, not just point estimates
- Wide CIs = more caution needed in interpretation
- Narrow CIs = more confidence in estimates

### 3. Compare Appropriately

- Use per capita rates when comparing countries
- Consider demographic differences (age structure, healthcare systems)
- Don't over-interpret small differences when CIs overlap

### 4. Think Clinically and Practically

- Statistical significance ≠ clinical importance
- Consider magnitude: Is a difference large enough to matter?
- Think about feasibility: Can identified groups be targeted effectively?

### 5. Document Your Exploration

- Take notes on patterns observed
- Save visualizations as you go
- Export data for verification and further analysis

### 6. Verify Findings

- Cross-check between tabs (e.g., prevalence vs. burden)
- Look for consistency across metrics
- Consider biological plausibility of patterns
